A Mechanised Library for Generic (Co)Programming

### 2024-03-01 paper/paper.tex
Algebras and Initial Algebras

\section{Algebras, Coalgebras and Hylomorphisms}
Recursion schemes provide an abstract way to consume and generate data.  The
most simple and well-known recursion scheme is the \emph{fold} or
\emph{catamorphism} which consumes data. Its dual is the \emph{unfold} or
\emph{anamorphisms} which produces data.

For example, assuming the type of lists over a type \coq{A},
namely \coq{List A}, and given an
algebra \coq{alg : unit + A * List A -> List A} a
catamorphism \coq{cata alg : List A -> B} destructs a list into its head and
tail using a destructor \coq{InOp : List A -> unit + A * List A}. This latter is
actually an isomorphim with its inverse
\begin{minted}{coq}
In : unit + A * List A -> List A
\end{minted}

defined by the maps \coq{Empty : unit -> List A}
and \coq{Cons : A * List A -> List A}.

After destructing the input the catamorphism is applying itsefl recursively.
This is done using functoriality of the shape of the data, which is a map of
type
\begin{minted}{coq}
unit + A * List A -> unit + A * List B
\end{minted}
Once the recursion as returned it result it applies the algebra \coq{alg} which
composes together the results of the recursion yielding the result of
type \coq{B}.

More abstractly, assuming a base functor \coq{F} a catamorphism is the unique
solution to the equation
\begin{minted}{coq}
cata alg = alg . fmap (cata alg) . InOp
\end{minted}

@@

% C
Hylomorphism are more general recursion schemes capturing the
\emph{divide-and-conquer} idea. In short, the input is destructured by its not
structurally. In the example, we gave in the introduction the destructuring
program is given by \coq{divide} which is not an isomorphism.  The question the
hylomorphism satisfies is more general than the catamorphism in that the
destructing map \coq{InOp} is generalised to a \emph{recursive
coalgebra}~\cite{AdamekMM19,CaprettaUV04}, that is intuitively a coalgebra that
can be applied only a finite amount of times.  In fact, in the case
of \coq{divide} this is a recursive coalgebra because the size of the list is
getting smaller.

With this in mind we know proceed onto describing how to formalise hylomorphisms
in Coq.

\subsection{Algebras}
Algebras on container extensions are defined as expected. The carrier of the
algebra is \coq{A}, and the operations of the algebra are morphisms of the
following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
Given a container extension \coq{F}, we define \emph{two}
(isomorphic) initial algebras. We will show the need for the second initial
algebra in the following section. Our fist initial algebra is defined as the
straightforward least fixpoint of the container extension \coq{F}:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words,
we would need to provide an equivalence for \coq{LFix}, but the
only way to do this is to define such equivalence simultaneously to the
definition for  \coq{LFix}. Such definition would be rejected by Coq.
There are other approaches that can circumvent this problem, by separating
the definition of least fixed points from the proofs that the functions that
get the contained elements are proper morphisms. But this leads to longer,
unnecessarily complicated proofs.  Furthermore, this extra layer interferes
with code extraction, in that it requires destructing/constructing subsequent
\coq{LFix} and  \coq{App}, and Coq may not always able to inline them.

@@

\subsection{(Recursive) Coalgebras}

@@

\subsection{Recursive Hylomorphisms}

>>>>

\section{Algebras, Coalgebras and Hylomorphisms}
Recursion schemes provide an abstract way to consume and generate data. We know
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise recursive coalgebras (Section~\ref{sec:rec-coalgebras}) and then we
put together these notions to formalise hylomorphisms (Section~\ref{sec:hylo}).

\subsection{Algebras}
\label{sec:algebras}
The most simple and well-known recursion scheme is the \emph{fold} or
\emph{catamorphism} which consumes data. Its dual is the \emph{unfold} or
\emph{anamorphisms} which produces data.

A fold consumes elements of an inductive data type. These can be thought as
least ``sets'' generated by some constructors. For example, the type of lists
over a type \coq{A}, namely \coq{List A} is the ``least set'' generated by the
constructors \coq{Empty : unit -> List A} and \coq{Cons : A * List A -> List A}.
Together, \coq{Empty} and \coq{Cons} form an \emph{algebra} of
type \coq{In : unit + A * List A -> List A}. In particular, this algebra is an
isomorphism with the obvious inverse \coq{InOp : List A -> unit + A * List A}
returning \coq{tt} is the list is empty and the head and tail of the list
otherwise. This type of algebra is called \emph{initial algebra}.  This is
because for for any other algebra \coq{alg : unit + A * List B -> B} there
exists a unique map which we call \coq{cata alg : List A -> B} which destructs a
list into its head and tail using a
destructor \coq{InOp : List A -> unit + A * List A}, applies itself recursively,
composes together the results of the recursion using the algebra \coq{alg}.

To generalise recursion schemes we abstract the shape of the data using functors.
Assuming a base functor \coq{F} a catamorphism is then the unique
solution to the equation
\begin{minted}{coq}
cata alg = alg . fmap (cata alg) . InOp
\end{minted}
which does precisely what we described in the previous paragraph.

@@

Given a type `\coq{A}` and a container \coq{F} an algebra for \coq{F} over the
carrier `\coq{A}` is a map of the following type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
When \coq{A} is the least fixed-point of \coq{F} this map is an isomorphism and
it is called the \emph{initial algebra} and denotes the inductive data type shaped by the functor represented by \coq{F}:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
This definition highlights why we cannot define equality of the positions and
the contained elements generically (recall Section~\ref{sec:TODO}): the
contained types may not have (yet) an associated equivalence. In other words, we
would need to provide an equivalence for \coq{LFix}, but the only way to do this
is to define such equivalence simultaneously to the definition for  \coq{LFix}.
Such definition would be rejected by Coq. There are other approaches that can
circumvent this problem, by separating the definition of least fixed points from
the proofs that the functions that get the contained elements are proper
morphisms. But this leads to longer, unnecessarily complicated proofs.
Furthermore, this extra layer interferes with code extraction, in that it
requires destructing/constructing subsequent \coq{LFix} and \coq{App}, and Coq
may not always able to inline them.

@@

\subsection{(Recursive) Coalgebras}
\label{sec:rec-coalgebras}

Hylomorphism are more general recursion schemes capturing the
\emph{divide-and-conquer} idea. In short, the input is destructured by its not
structurally. In the example, we gave in the introduction the destructuring
program is given by \coq{divide} which is not an isomorphism.  The question the
hylomorphism satisfies is more general than the catamorphism in that the
destructing map \coq{InOp} is generalised to a \emph{recursive
coalgebra}~\cite{AdamekMM19,CaprettaUV04}, that is intuitively a coalgebra that
can be applied only a finite amount of times.  In fact, in the case
of \coq{divide} this is a recursive coalgebra because the size of the list is
getting smaller.

@@

\subsection{Hylomorphisms}
\label{sec:hylo}
